Individual heterogeneity in the functional topography of the medial prefrontal cortex

### 2019-03-15 Manuscript.Rmd
complete restructuring of the tSNR section

@@

### *Potential Noise Effects*

*NOTE: this part seems less relevant now that we are thresholding the partitions. Might be better to add to sliding window section*

``` {r Noise Effects: tSNR Comparison, echo=FALSE, message = FALSE, cache = TRUE}
# Compare SNR between PCC and mPFC per subject after thresholding (this filtering is relevant here because we aren't using ARI anymore)
tempComm_PCC <- lapply(tempComm_PCC, filter, FiedlerVec != 0)
tempComm_PFC <- lapply(tempComm_PFC, filter, FiedlerVec != 0) 
SNR_PCCvsPFC_perm <- sapply(seq_along(tempComm_PCC), function(x) {permute(tempComm_PCC[[x]]$varCoeff, tempComm_PFC[[x]]$varCoeff)$Pval})
SNR_PCCvsPFC_perm <- p.adjust(SNR_PCCvsPFC_perm, method = "BY")
SNR_PCCvsPFC_ES <- sapply(seq_along(tempComm_PCC), function(x) {cohenD(tempComm_PCC[[x]]$varCoeff, tempComm_PFC[[x]]$varCoeff)})
prcntSig <- mean(SNR_PCCvsPFC_perm < 0.05) * 100

# Get the mean SNR for each region and compare overall levels 
SNR_meanPCC <- sapply(tempComm_PCC, function(data) {mean(data$varCoeff)})
SNR_meanPFC <- sapply(tempComm_PFC, function(data) {mean(data$varCoeff)})
SNR_meanComp <- list(Values = data.frame(Region = c(rep("PCC", nSubj), rep("mPFC", nSubj)), SNR = c(SNR_meanPCC, SNR_meanPFC)),
                     Comparison = permute(SNR_meanPCC, SNR_meanPFC)$Pval,
                     ES = cohenD(SNR_meanPCC, SNR_meanPFC))
rm(SNR_meanPCC, SNR_meanPFC)
```

Even though our findings point to a stable representation of each subject's unique DMN organization, it is still possible that they were driven by considerable levels of signal noise often found in mPFC (its ventral segment in particular). To account for any residual noise, we computed the temporal signal to noise ratio (tSNR) for each vertex in each subject's minimally preprocessed fMRI data (i.e. mean of each vertex's time series over its standard deviation). Note that we did this for the vertices that survived the threshold. The tSNR of all vertices were averaged per region and subject (Figure 11), which showed that mPFC (mean = `r round(mean(SNR_meanComp$Values$SNR[SNR_meanComp$Values$Region=="mPFC"]), digits = 2)`; SD = `r round(sd(SNR_meanComp$Values$SNR[SNR_meanComp$Values$Region=="mPFC"]), digits = 2)`) and PCC (mean = `r round(mean(SNR_meanComp$Values$SNR[SNR_meanComp$Values$Region=="PCC"]), digits = 2)`; SD = `r round(sd(SNR_meanComp$Values$SNR[SNR_meanComp$Values$Region=="PCC"]), digits = 2)`) had similar tSNR (permutation, *p* = `r SNR_meanComp$Comparison`, Cohen's D = `r round(SNR_meanComp$ES, digits = 2)`). To more precisely examine this difference, we performed subject-wise comparisons of the tSNR values of all vertices between PCC and mPFC (5000 iterations each). About `r round(prcntSig, digits = 2)` percent of the comparisons proved significant at alpha = 0.05 (FDR corrected), but these differences had low effect sizes (mean Cohen's D = `r round(mean(SNR_PCCvsPFC_ES), digits = 2)`). These findings further confirm that the idiosyncrasies observed in these areas (and in mPFC in particular) are not a function of measurement noise.

### new stuff
# append the Rsquared values onto the current summaries
# unfortunately this only counts on the DMN community
appendVar <- lapply(seq(nSubj), function(subj) {left_join(Summaries[[subj]], Variance[[subj]], by = "Vertex")})

# correlate tSNR with Rsquared before and after thresholds
snrR2_cors <- list()

snrR2_cors$pre_thresh <- sapply(appendVar, function(data) {
  data %>% 
    filter(!is.na(Rsquared)) %>%
    select(varCoeff, Rsquared) %>%
    cor(.$varCoeff)
})[2, ]

snrR2_cors$post_thresh <- sapply(appendVar, function(data) {
  data %>% 
    filter(!is.na(Rsquared), FiedlerBinary != 0.5) %>%
    select(varCoeff, Rsquared) %>%
    cor(.$varCoeff)
})[2, ]

# next get the mean tSNR of PCC and mPFC before/after
snrR2_means <- list()
snrR2_means <- sapply(appendVar, function(data) {
  data %>% 
    filter(!is.na(Rsquared)) %>%
    mutate(region = ifelse(z > 0, "mPFC", "PCC")) %>%
    group_by(region) %>% 
    summarize(mean_tsnr = mean(varCoeff)) %>%
    ungroup()
})

### end new stuff

### Grab some of the other measures from the overall summary
for (i in seq(nSubj)) {
  tempIndex <- Summaries[[i]]$Vertex %in% Variance[[i]]$Vertex
  Variance[[i]]$SubjID <- rep(subjList[i], nrow(Variance[[i]]))
  Variance[[i]]$FiedlerVec <- Summaries[[i]]$FiedlerVec[tempIndex]
  Variance[[i]]$Flexibility <- Summaries[[i]]$slideFlexibility[tempIndex]
  Variance[[i]]$propDMN <- Summaries[[i]]$slidePropDMN[tempIndex]
  Variance[[i]]$varCoeff <- Summaries[[i]]$varCoeff[tempIndex]
  Variance[[i]]$Region <- ifelse(Variance[[i]]$tempY < 0, "PCC", "mPFC")
}

# let's put everyone's data together and filter out according to the threshold
allVariance <- do.call(rbind, Variance) %>%
                  filter(FiedlerVec != 0)

# get the mean and SE of the R-squared values per subject/region
allVariance_sumarized <- aggregate(allVariance$Rsquared ~ allVariance$SubjID + allVariance$Region, FUN = 'mean')
colnames(allVariance_sumarized) <- list("SubjID", "Region", "meanRsquared")
allVariance_sumarized$SEM <- aggregate(allVariance$Rsquared ~ allVariance$SubjID + allVariance$Region, FUN = function(data) {sd(data)/sqrt(length(data))})[ , 3]

# Create a new data frame with values restricted to mPFC (for plotting and further stats)
mPFCVariance <- allVariance[allVariance$Region == "mPFC", ]

# Get the correlations between Z-axis, R2, and tSNR for every subject (to take mean and SD in text)
# I just came up with this, so the code should be made more parsimonious in the future (like, why bind and then split...?)
tempSubj <- split(mPFCVariance, f = mPFCVariance$SubjID)
# Get the correlation coefficients
noiseCors <- data.frame(R2vtSNR = sapply(tempSubj, function(data) {cor(data$Rsquared, data$varCoeff)}),
                        ZvtSNR = sapply(tempSubj, function(data) {cor(data$tempZ, data$varCoeff)}),
                        ZvR2 = sapply(tempSubj, function(data) {cor(data$tempZ, data$Rsquared)}))
rm(tempSubj)

# plot
# There is a strong linear relationship between the Fiedler vector and the R-squared values
# These match perfectly on the brain when inspected qualitatively
v1 <- ggplot(data = allVariance) + 
  geom_smooth(aes(FiedlerVec, Rsquared, group = SubjID), fill = "gray70", color = "gray30", show.legend = F) + 
  labs(x = "Fiedler Vector Values", y = "R-Squared Values") + 
  theme_classic()

# The lower the R-squared, the more likely it is for nodes to change affiliation throughout time
v2 <- qplot(data = allVariance, Rsquared, Flexibility, color = SubjID, fill = SubjID, geom = "smooth") + theme_classic()

# And of course (given the relationship with the Fiedler vector), there's a sigmoidal relationship between R-squared vals and proportion of DMN affiliation
# only showing the upper side of the sigmoid here, as these values are restricted to the DMN community
v3 <- qplot(data = allVariance, Rsquared, propDMN, color = SubjID, fill = SubjID, geom = "smooth") + theme_classic()

# Finally, it's worth comparing the R-squared values between PCC and mPFC
# The intuition is that the variance explained by the community might be lower for mPFC given its noise
# And the answer is: mPFC yields significantly lower R-squares per subject and overall, as expected.
# To formally compare this (and given the binomial distribution of R-squares in mPFC), I report the Cohen's D per subject for PCC vs mPFC
v4 <- ggplot(data = allVariance_sumarized, aes(x = SubjID, y = meanRsquared, group = Region, color = Region)) +
  geom_point(show.legend = F) +
  geom_line(show.legend = F) +
  geom_errorbar(aes(ymin = meanRsquared - SEM, ymax = meanRsquared + SEM)) +
  scale_color_manual(values = Cols[seq(2)]) +
  ylim(0,0.5) +
  labs(x = "Subject", y = "Mean R-squared") +
  theme_classic() +
  theme(axis.text.x = element_blank())

v5 <- qplot(data = allVariance, Region, Rsquared, fill = Region, geom = "boxplot") + theme_classic()
Rsq_PCCvsPFC_Wilcox <- sapply(Variance, function(data) {wilcox.test(data$Rsquared[data$Region == "PCC"], data$Rsquared[data$Region == "mPFC"])$p.value})
Rsq_PCCvsPFC_Wilcox <- p.adjust(Rsq_PCCvsPFC_Wilcox, method = "BY")
prcntSig <- (sum(Rsq_PCCvsPFC_Wilcox < 0.05) / nSubj) * 100
Rsq_PCCvsPFC_ES <- sapply(Variance, function(data) {cohenD(data$Rsquared[data$Region == "PCC"], data$Rsquared[data$Region == "mPFC"])})

# This begs the question of a "gradient of noise" along the z-axis on mPFC
# Not really, but the coefficient of variation derived from minimally preprocessed data does show a strong trend
# Namely, the more ventral, the lower the SNR
# v6 <- ggplot(data = allVariance, aes(Rsquared, tempZ, fill = Region)) +
#         geom_point(color = "black", pch = 21, alpha = 0.1) + 
#         scale_fill_manual(values = Cols[seq(2)]) +
#         theme_classic()
# v7 <- ggplot(data = allVariance, aes(varCoeff, tempZ, fill = Region)) +
#         geom_point(color = "black", pch = 21) + 
#         scale_fill_manual(values = Cols[seq(2)]) +
#         theme_classic()

#v1

To further interrogate the potential effect of noise in our results, we computed the variance explained by activity from each DMN vertex on the mean time series of the DMN community. If the signal quality issues persisted after preprocessing the data, we expected that noisy vertices would have the least explicative power on the activity of the DMN community as a whole. This also implied lower $R^2$ values for mPFC. Indeed, we found that for each subject, $R^2$ values tended to be higher for PCC (`r prcntSig` percent of Wilcoxon ranksum tests across subjects had *p* values < 0.001; mean Cohen's D = `r round(mean(Rsq_PCCvsPFC_ES), digits = 2)`). These results suggest that, overall, vertices in mPFC tended to be less in synchrony with the activity of the DMN community as a whole. Furthermore, the level of tSNR showed a moderate correlation with $R^2$ values across participants (mean *r* = `r round(mean(noiseCors$R2vtSNR), digits = 2)`; SD = `r round(sd(noiseCors$R2vtSNR), digits = 2)`), meaning that noisier vertices were less in synchrony with the general activity of the DMN.

Given this trend, we asked whether there is a "gradient of noise" along the z-axis of mPFC, such that more ventral vertices would both be noisier and explain the least amount of DMN variance. Here we found that while $R^2$ are somewhat correlated with dorsal-ventral placement across subjects (mean *r* = `r round(mean(noiseCors$ZvR2), digits = 2)`; SD = `r round(sd(noiseCors$ZvR2), digits = 2)`), there is no clear relationship with the z-axis across subjects (Figure 12). Specifically, low $R^2$ are evenly distributed along this axis, denoting a degree of orthogonality between these features. Conversely, tSNR showed a strong linear relationship with the position of the vertex along this axis (mean *r* = `r round(mean(noiseCors$ZvtSNR), digits = 2)`; SD = `r round(sd(noiseCors$ZvtSNR), digits = 2)`). These results contextualize the degree of synchrony of mPFC vertices in terms of potential noise. While still an important factor, it is unlikely that the degree of within-subject pattern stability observed here could be due mostly to highly structured mPFC noise.

@@

``` {r Noise Effects: tSNR Comparison (Fig. 12), fig.align="center", fig.width=3, fig.height=3, fig.cap = "WEIRD: Mean tSNR for mPFC and PCC across participants. This measure was derived from the minimally preprocessed fMRI data from the HCP, and thus shows the base level of reltive noise in each region.", echo=FALSE, message = FALSE}
# Plot the mPFC SNR vs PCC SNR for each subject (one figure containing all subjects)
snrmeanComp_plot <- ggplot(data = SNR_meanComp$Values, aes(Region, SNR, fill = Region)) +
  geom_boxplot(show.legend = F) +
  scale_fill_manual(values = Cols[seq(2)]) +
  labs(x = "", y = "tSNR") +
  theme(text = element_text(size=35)) +
  theme_classic()
snrmeanComp_plot
```

<!--
``` {r Noise Effects: Mean R-sqd mPFC and PCC (Fig. 13), fig.align="center", fig.width=6, fig.height=4, fig.cap = "Differences in mean R-squared values for each region per participant. The figure shows that PCC vertices tend to be more coupled with the overall dynamics of DMN across subjects.", echo=FALSE, message = FALSE}
#v4
```
-->

``` {r Noise Effects: R-sqd vs tSNR (Fig. 13), fig.align="center", fig.width=5, fig.height=3, fig.cap = "Vertex tSNR values across all subjects are correlated with their position along the z-axis, with vmPFC showing the lowest tSNR (left). However, R-square values do not show this strong relationship (right), suggesting that the coupling of each vertex is not necessarily influenced by signal noise.", echo=FALSE, message = FALSE}
# grid.arrange(v6, v7, ncol = 2)
# par(mfrow = c(1,2))
# smoothScatter(allVariance[allVariance$Region == "mPFC",]$Rsquared, allVariance[allVariance$Region == "mPFC",]$tempZ,
#               xlab = "R-squared Value",
#               ylab = "Z-Coordinate")
# 
# smoothScatter(allVariance[allVariance$Region == "mPFC",]$varCoeff, allVariance[allVariance$Region == "mPFC",]$tempZ,
#               xlab = "Coefficient of Variation (SNR)",
#               ylab = "")

# ALTERNATIVE IDEA FOR A FIGURE: BRAIN MAPS WITH THESE VALUES. THAT SAID, R2 MAPS LOOK THE SAME AS FV ONES, SO...
s1 <- ggplot(data = mPFCVariance, aes(varCoeff, tempZ)) +
  geom_bin2d(bins = 250, show.legend = F) +
  scale_fill_continuous(low = "grey30", high = "grey70") +
  geom_density2d(color = Cols[2]) +
  labs(x = "tSNR", y = "Z-Coordinate") +
  theme_classic()

s2 <- ggplot(data = mPFCVariance, aes(Rsquared, tempZ)) +
  geom_bin2d(bins = 250, show.legend = F) +
  scale_fill_continuous(low = "grey30", high = "grey70") +
  geom_density2d(color = Cols[2]) +
  labs(x = "R-squared value", y = "Z-Coordinate") +
  xlim(c(0,1)) +
  theme_classic()

grid.arrange(s1, s2, ncol = 2)
```

>>>>

## append the Rsquared values onto the current summaries
# unfortunately this only counts on the DMN community
appendVar <- lapply(seq(nSubj), function(subj) {left_join(Summaries[[subj]], Variance[[subj]], by = "Vertex")})

## correlate tSNR with Rsquared before and after thresholds
snrR2_cors <- list()

snrR2_cors$pre_thresh <- lapply(appendVar, function(data) {
  data %>% 
    filter(!is.na(Rsquared)) %>%
    mutate(region = ifelse(y > 0, "mPFC", "PCC")) %>%
    group_by(region) %>%
    dplyr::summarise(correlation = cor(varCoeff, Rsquared)) %>%
    ungroup()
}) %>% reduce(left_join, by = "region")

snrR2_cors$post_thresh <- lapply(appendVar, function(data) {
  data %>% 
    filter(!is.na(Rsquared), FiedlerBinary != 0.5) %>%
    mutate(region = ifelse(y > 0, "mPFC", "PCC")) %>%
    group_by(region) %>%
    dplyr::summarise(correlation = cor(varCoeff, Rsquared)) %>%
    ungroup()
}) %>% reduce(left_join, by = "region")

## next get the mean tSNR of PCC and mPFC before/after thresholding
# the first piece of code gets each subject's mean tSNR per region, and reduces to a single matrix
# the second one is similar, but filtering out vertices that didn't meet the threshold
# the third one applies summary stats to each list to create a summary table
snrR2_means <- list()

snrR2_means$pre_thresh <- lapply(appendVar, function(data) {
  data %>% 
    mutate(region = ifelse(y > 0, "mPFC", "PCC")) %>%
    group_by(region) %>% 
    dplyr::summarise(mean_tsnr = mean(varCoeff)) %>%
    ungroup()
}) %>% reduce(left_join, by = "region")

snrR2_means$post_thresh <- lapply(appendVar, function(data) {
  data %>% 
    filter(FiedlerBinary != 0.5) %>%
    mutate(region = ifelse(y > 0, "mPFC", "PCC")) %>%
    group_by(region) %>% 
    dplyr::summarise(mean_tsnr = mean(varCoeff)) %>%
    ungroup()
}) %>% reduce(left_join, by = "region")

# and compute summary stats on each 
snrR2_means$overall <- sapply(snrR2_means, function(x) {
  c(rowMeans(x[, -1]), apply(x[, -1], 1, sd))
})
rownames(snrR2_means$overall) <- c("mPFC_mean", "PCC_mean", "mPFC_SD", "PCC_SD")

# permutation pre vs post
tsnrPermute_mPFC <- permute(as.numeric(snrR2_means$pre_thresh[1, -1]), as.numeric(snrR2_means$post_thresh[1, -1]), paired = T)
tsnrPermute_PCC <- permute(as.numeric(snrR2_means$pre_thresh[2, -1]), as.numeric(snrR2_means$post_thresh[2, -1]), paired = T)

# effect size pre vs post
tsnrCohen_mPFC <- CohenD(as.numeric(snrR2_means$post_thresh[1, -1]), as.numeric(snrR2_means$pre_thresh[1, -1]))
tsnrCohen_PCC <- CohenD(as.numeric(snrR2_means$post_thresh[2, -1]), as.numeric(snrR2_means$pre_thresh[2, -1]))

# ##### old
# 
# # Compare SNR between PCC and mPFC per subject after thresholding (this filtering is relevant here because we aren't using ARI anymore)
# tempComm_PCC <- lapply(tempComm_PCC, filter, FiedlerVec != 0)
# tempComm_PFC <- lapply(tempComm_PFC, filter, FiedlerVec != 0) 
# SNR_PCCvsPFC_perm <- sapply(seq_along(tempComm_PCC), function(x) {permute(tempComm_PCC[[x]]$varCoeff, tempComm_PFC[[x]]$varCoeff)$Pval})
# SNR_PCCvsPFC_perm <- p.adjust(SNR_PCCvsPFC_perm, method = "BY")
# SNR_PCCvsPFC_ES <- sapply(seq_along(tempComm_PCC), function(x) {cohenD(tempComm_PCC[[x]]$varCoeff, tempComm_PFC[[x]]$varCoeff)})
# prcntSig <- mean(SNR_PCCvsPFC_perm < 0.05) * 100
# 
# # Get the mean SNR for each region and compare overall levels 
# SNR_meanPCC <- sapply(tempComm_PCC, function(data) {mean(data$varCoeff)})
# SNR_meanPFC <- sapply(tempComm_PFC, function(data) {mean(data$varCoeff)})
# SNR_meanComp <- list(Values = data.frame(Region = c(rep("PCC", nSubj), rep("mPFC", nSubj)), SNR = c(SNR_meanPCC, SNR_meanPFC)),
#                      Comparison = permute(SNR_meanPCC, SNR_meanPFC)$Pval,
#                      ES = cohenD(SNR_meanPCC, SNR_meanPFC))
# rm(SNR_meanPCC, SNR_meanPFC)

To discard the possibility that the higher inter-subject variability in mPFC was driven by retaining vertices with lower signal quality, we compared the mean temporal signal to noise ratio (tSNR) for each region pre and post thresholding. We used subject time series that were not demeaned, but were otherwise equivalent to the data originally used. Paired permutation analyses (5000 iterations) showed that the mean tSNR in mPFC actually significantly increased post-thresholding across subjects (Pre: mean = `r snrR2_means$overall[1, 1]`, SD = `r snrR2_means$overall[3, 1]`; Post: mean = `r snrR2_means$overall[1, 2]`, SD = `r snrR2_means$overall[3, 2]`; paired permutation p-value = `r tsnrPermute_mPFC$Pval`, Cohen's D = `r tsnrCohen_mPFC[1]`), whereas the signal quality in PCC was preserved (Pre: mean = `r snrR2_means$overall[2, 1]`, SD = `r snrR2_means$overall[4, 1]`; Post: mean = `r snrR2_means$overall[2, 2]`, SD = `r snrR2_means$overall[4, 2]`; paired permutation p-value = `r tsnrPermute_PCC$Pval`, Cohen's D = `r tsnrCohen_PCC[1]`).

@@

@@
